---
title: "Practical One"
author: "Karabo Tigedi"
format: 
  html: 
    toc: true
    theme: 
        dark: darkly
    toc-depth: 3
    toc-location: left
    toc-title: "On this page"
    code-links:
      - text: Github link source code
        href: https://data.nasa.gov/
    css: csss
    code-overflow: wrap
   # code-tools: 
      #  source: true
---

# Question 1

Airquality (available in R) was utilised for this question. Here is a sneak peak of the dataset:

```{r}
#| echo: false
quality_data = airquality
head(quality_data)
```

The data consists of Ozone, Solar.R, Wind, Temp, Month, and Day columns.

```{r}
#| code-fold: true
quality_data = airquality
head(quality_data[(is.na(quality_data$Ozone) | is.na(quality_data$Temp)),])
tail(quality_data[(is.na(quality_data$Ozone) | is.na(quality_data$Temp)),])
```

# Question 2

The following statistics: mean, sd, min, and max were derived for each of temperature and ozone level, and they are as follows:

```{r}
#| code-fold: true
temp_info = data.frame("Mean" = mean(quality_data$Temp),
                       "Sd" = sd(quality_data$Temp), 
                       "Min" = min(quality_data$Temp), 
                       "Max" = max(quality_data$Temp))
temp_info
```

However, the Ozone column has a few rows with missing data(NAs):

```{r}
#| echo: false
cat("[OUTPUT] Number of missing values for Ozone column:","\t", sum(is.na(quality_data$Ozone)))
```

So a slight adjustment was made to the code to find the missing values. Please select the *Code* button below to see the slight adjustment. However, here are the statistics:

```{r}
#| code-fold: true
Ozone_info = data.frame("Mean" = mean(quality_data$Ozone, na.rm = T),
                       "Sd" = sd(quality_data$Ozone, na.rm = T), 
                       "Min" = min(quality_data$Ozone, na.rm = T), 
                       "Max" = max(quality_data$Ozone, na.rm = T))

Ozone_info
```

# Question 3

Now we move to a different dataset called *Cars* available in R. Here is a sneak peak of the data:

```{r}
#| echo: false
cars_data = cars
head(cars_data)
```

In this section we wanted to, from first principle or Using matrix calculations, find linear regression beta estimates, t-statistics, r squared, standard errors, etc, and fortunately we found the following:

```{r}
#| echo: false
x = cbind(1,cars_data$speed)
y = cars_data$dist
beta_ests = (solve(t(x)%*%x))%*%t(x)%*%y
cars_data = cars
y_hat = x%*%beta_ests
rss = (y - y_hat)^2 |> sum() # Residual Sum of Squares
tss = (y - mean(y))^2 |> sum()   # Total Sum of Squares
r_squared = (tss - rss)/ tss
n = length(y)  # Number of observations
p = ncol(x)    # Number of pars including intercept
residual_variance = rss / (n -p)  # Estimate of residual variance
covariance_matrix = residual_variance*solve(t(x) %*% x)
standard_errors = sqrt(diag(covariance_matrix))
t_stats = beta_ests / standard_errors
p_values = 2 *(1 - pt(abs(t_stats), df = n -p ))
list(
  Betas_estimates = beta_ests,
  RSS = rss,
  R_squared = r_squared,
  Standard_errors = standard_errors,
  P_values = p_values,
  T_statistics = t_stats
)
```

# Question 4

Now we use lm() function in R to fit the same data and evaluate whether the estimates from the model are consistent with our first principles or not:

```{r}
#| echo: false
# Use lm directly
model = lm(dist ~ speed, data = cars_data)
summary(model)
```

As you can see, the results are quite the same. Thank God!! We are so smart! Interpretation: As speed increases by one unit, distance to stop increases by 3.932 (holding everything else constant).
